---
title: "Getting started with compfit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with compfit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE      # shown, not run: most chunks want a fitted model or Julia
)
```

You describe a compartmental (SIR-type) ODE model in three small spreadsheets,
and `compfit` fits it, by maximum likelihood or Bayesian sampling. It then helps
you analyse the result: posterior summaries, predictive plots, global (Sobol) and
local sensitivity, practical identifiability, and reproducible code export. You
write down the model; the package handles the ODE build, the loss or likelihood,
the solver, and the plots.

This guide walks through that workflow and the sheet grammar it uses. For a
worked example that runs live, with real output and figures, see the companion
article *"Fitting an SIR model with compfit"*.

## At a glance

The basic workflow is three calls: load, fit, plot.

```{r}
library(compfit)

# A scenario is a folder of three sheets: model, data, and optional dummy series.
sc <- load_scenario("path/to/MyScenario",
                    combined_file = "dataCombined.csv",
                    dummy_file    = "dataDummy.csv",
                    params_file   = "modelParams.csv")

# Fit by maximum likelihood. backend = "r" runs entirely in R, with no Julia.
fit <- fitCompartmentalModel(sc$modelParams, sc$dataCombined,
                             method = "lbfgsb",
                             solver = solver_control(backend = "r"))
summary(fit)

# Inspect the fit (and any dummy series) as a grid of per-stream panels.
plot_fit(fit, data_dummy = sc$dataDummy)$grid
```

The rest of this guide covers each step.

## Choosing a backend

Solving the ODE, and for Bayesian fits drawing the posterior, can run on any of
three backends. Pick one with `solver_control(backend = ...)`:

| backend | engine | needs | good for |
|----|----|----|----|
| `"julia"` (default) | Julia `OrdinaryDiffEq` + Turing NUTS | a Julia session | fastest solves; gradient-based Bayesian sampling with full feature support |
| `"r"` | `deSolve` + gradient-free MCMC (`BayesianTools`) | nothing extra | Julia-free fitting; small models, quick checks, teaching |
| `"stan"` | Stan NUTS via `rstan` | a C++ toolchain | gradient-based Bayesian sampling, no Julia; best for censored counts |

Maximum-likelihood fits give the same answer on every backend, just faster on
Julia. For Bayesian fits the differences matter:

- `"julia"` is the main path: fast NUTS, every family and every data-cell type.
- `"r"` uses a gradient-free sampler (`BayesianTools` DEzs), seeded from a quick
  MLE fit. It is much slower and mixes far worse, so it is best for a small model
  or a sanity check. For real inference use Julia or Stan.
- `"stan"` gives gradient-based NUTS with no Julia. All three backends now score
  every data-cell type (observed, censored, interval `[A,B]`/`[A,B]~s`,
  asymmetric) and all four families. For *censored count* data Stan is actually
  the strongest option (see the CDF note below).

Every data-cell type now works on all three Bayesian backends. The one remaining
backend nuance is that Julia cannot score a censored/interval/asymmetric cell on
a **discrete** family (see the CDF note); Stan and R can.

Everything downstream (`posterior_report()`, `plot_prior_posterior()`, and so on)
works the same whichever backend you pick.

The Julia backend needs a one-time setup per session. The first call also installs
the required Julia packages (OrdinaryDiffEq, Turing, Distributions), which takes a
few minutes; later calls are quick:

```{r}
setup_julia()
```

The bridge also initialises lazily, so `setup_julia()` mainly just does the slow
part up front. The R and Stan backends need no Julia at all.

### Choosing the Julia ODE solver

On the Julia backend, `solver_control(solver = ...)` takes any `OrdinaryDiffEq`
solver, written as the Julia constructor and passed straight through to the
solve. The default auto-switches between a non-stiff and a stiff method and is a
good fit for almost every compartmental model, so you rarely need to change it.
When you do:

| `solver` | Kind | Use when |
|----|----|----|
| `"AutoTsit5(Rosenbrock23())"` *(default)* | auto-switching | general purpose; leave it here unless a fit misbehaves |
| `"Tsit5()"` | non-stiff | plain, fast, smooth dynamics |
| `"Vern7()"`, `"Vern9()"` | non-stiff, high accuracy | tight tolerances |
| `"Rosenbrock23()"`, `"Rodas5()"`, `"Rodas4()"` | stiff (Rosenbrock) | stiff systems (fast/slow rates) |
| `"KenCarp4()"`, `"TRBDF2()"` | stiff (implicit) | larger stiff systems |
| `"AutoVern7(Rodas5())"` | auto-switching | like the default but higher accuracy |

```{r}
fit <- fitCompartmentalModel(sc$modelParams, sc$dataCombined, method = "lbfgsb",
                             solver = solver_control(solver = "Rodas5()"))
```

The full menu is in the
[DifferentialEquations.jl solver guide](https://docs.sciml.ai/DiffEqDocs/stable/solvers/ode_solve/).
This choice is Julia-only: the R backend uses deSolve's auto-switching `lsoda`,
and the Stan backend uses Stan's `ode_rk45` (or `ode_bdf` via
`solver_control(solver = "bdf")`).

## The scenario sheets

A *scenario* is one folder holding the input files:

```
MyScenario/
  modelParams.xlsx    # the model: compartments, parameters, rates, time grid
  dataCombined.xlsx   # the data: one row per observed stream (optional; omit to simulate)
  dataDummy.xlsx      # optional derived series to overlay
```

`load_scenario()` reads all three. `sc` then holds `sc$modelParams`,
`sc$dataCombined`, `sc$dataDummy`, and a `sc$plot_path` for output. Formats
`.csv`, `.xlsx`, `.tsv`, `.rds`, `.parquet` and more are detected automatically.

```{r}
sc <- load_scenario(
  "path/to/MyScenario",
  combined_file = "dataCombined.xlsx",
  dummy_file    = "dataDummy.xlsx",
  params_file   = "modelParams.xlsx"
)
```

### modelParams: the model

The model sheet uses column groups:

- **`States`**: the compartments, in order, with their initial values. You can
  name them freely, `S`, `I`, `R` or `X1`, `X2`, `X3`. The `States` column sets
  both the names and the order. Each entry is `*name=val` (fixed), `name=[lo,hi]`
  (fitted), or `*name=<expr>` (a parameter-dependent initial value, for example
  `*X2=init_inf_0*N0_0`).
- **`Parameters`**: `*name=val` is fixed. A fitted parameter is given by its
  **prior**, either the box form `name=[lo,hi]` (Uniform) or a named distribution
  (see the priors table below). A Bayesian fit uses the stated distribution as
  the prior. An MLE fit only needs a search box, so an unbounded prior (a plain
  `Normal`, say) is given a finite range from the distribution (`mean ± 4 sd`, or
  its natural support). Add a trailing `[lo,hi]` to set the MLE bounds exactly.
- **Start value** (optional): append `|value` to a box to set the optimiser's
  starting point, for example `beta=[0,3]|2.5`. Without it the search starts at
  the box midpoint. Use `|random` (or `|rand`) instead to draw the start
  uniformly from the box, for example `beta=[0,3]|random`; the draw is seeded by
  `optim_control(seed = )`, so it is reproducible, and `seed = NULL` gives a
  fresh start each run (handy for multi-start / escaping a poor local optimum).
  This drives the L-BFGS-B start (and the Bayesian MAP warm-start); `deoptim` and
  `hypercube` scan the whole box already, so a single start is moot there.
- **`Linear<j>` / `Quadratic<j>`**: the first- and second-order rate coefficients,
  one pair of columns per compartment. A blank cell counts as zero, and so does a
  whole column you leave out (that compartment simply gets no such terms).
- **`Constant`** (optional): a zeroth-order term added straight to a compartment's
  rate, one value per compartment (row `i` feeds `dX[i]`). Use it for an inflow
  that does not depend on any state, such as a constant birth or immigration rate,
  which cannot be written as a `Linear` (`*X[j]`) or `Quadratic` (`*X[i]*X[j]`)
  term. A cell may be a number, a parameter name, an expression of parameters, or a
  `$`-prefixed time-function (`$1000*exp(-0.1*time)`) -- the same grammar as a
  `Linear` coefficient, minus the `*X[j]`. A blank cell or an absent column is zero.
- **`Others`**: the time grid, with `startpoint`, `endpoint`, `partition`, and an
  optional `cutoff`. The axis is annual (integer years), and by convention the
  initial state sits at `startpoint - 1`.
- **`Functions`** (optional): time-varying helpers, for example
  `gamma_t<-gamma*(1+ramp*time)`.
- **`Conditions`** (optional): constraints turned into a fitting penalty, for
  example `beta>gamma`.
- **`_Level<j>`** (optional): which compartments share a mixing population (age,
  sex, region). The name after `_` is up to you (`_Level1`, `_Age1`, and so on),
  and cells take a compartment number or name. Leave it out and all compartments
  form a single population.

> **Reserved names.** A Bayesian fit uses `sigma` (the Gaussian or log-normal
> noise scale) and `phi`/`phi<i>` (negbin dispersion) as its own parameters, so a
> model parameter or state cannot be named `sigma` or `phi`. Using one raises a
> clear error on `method = "bayes"`; MLE is unaffected. Just rename it, for
> example an SEIR incubation rate as `sigma_ei`.
>
> Separately, a handful of names collide with variables the code generator emits
> and are rejected for **any** fit (parameters, states, and functions alike):
> Julia keywords and the solver internals `t`, `p`, `X`, `du`, `dX`, `parms`; the
> level populations `N1`, `N2`, … and their sum `total_pop`; and the per-term
> helpers `f<ij>`, `g<ij>`, `cst<i>`, `secOrd_<i>_<j>`. Also avoid declaring both a
> quantity and its `_0` alias (e.g. `beta` and `beta_0`). Note `N0` is fine (a
> common fixed initial-population parameter) -- only `N1` upward are taken.

The sheet is checked up front by `validate_modelParams()`, which reports entry
mistakes with a message naming the offending column and cell: a non-numeric fixed
value, a reversed or non-numeric box `[lo,hi]`, an out-of-range `Quadratic`
target, a typo in a coefficient, or a missing time-grid field. You can call it
yourself, or turn off the automatic check with
`options(compfit.validate = FALSE)`.

#### Priors

A fitted parameter or state can take any of these priors, written in the `States`
or `Parameters` cell:

| Prior | Sheet syntax | Support |
|----|----|----|
| Uniform | `[lo,hi]` | `[lo,hi]` |
| Normal | `Normal(mu,sd)` | real line |
| Log-normal | `LogNormal(mu,sd)` | positive |
| Beta | `Beta(a,b)` | `(0,1)` |
| Gamma | `Gamma(a,b)` | positive |
| Student-t | `StudentT(nu,mu,sigma)` | real line (heavy tails; `nu` = df) |

Every named distribution (all but the bare box) can be **truncated** by appending
a `[lo,hi]`, for example `Normal(0.4,0.1)[0,1]` or `LogNormal(-1,0.5)[0,5]`. The
`SIR_priors` and `SEIR_priors` example scenarios exercise all of them.

### dataCombined: the observations

One row per observed stream: a `Formula` (what the stream measures, for example
`X2` or `rho*X2`), an optional `Label` (its display name, which defaults to the
`Formula` when omitted or blank), an optional `Likelihood` family, and one column
per year.

`Weight` and `Average` are optional per-stream scaling columns for the MLE
objective, and they do not affect Bayesian fits. Omit them and `Weight` defaults
to 1 while `Average` is set to `1 / mean(observed)` per stream, so streams of
different magnitude count comparably. Include either column to override.

Each data cell is parsed as one of:

| Cell | Meaning |
|----|----|
| a number | observed value (`0` is a genuine zero) |
| `x` or blank | missing / unobserved |
| `<L` / `<=L` | left-censored: true value strictly / at-or-below `L` |
| `>L` / `>=L` | right-censored: true value strictly / at-or-above `L` |
| `[A,B]` | interval-censored: true value in `[A,B]`, flat within (`A < B`) |
| `[A,B]~s` | **soft** interval: plateau `[A,B]`, exponential shoulders of scale `s` (`~sl,su` for asymmetric shoulders) |
| `A->B` | asymmetric: recorded `A`, soft toward `B` (deviation `\|B-A\|`) |
| `A+` / `A-` | asymmetric: recorded `A`, soft up/down, deviation from the stream's `Likelihood` column |

The CDF-based cells (`<`, `<=`, `>`, `>=`, `[A,B]`, `A->B`, `A+`, `A-`) work for
every family on the R backend. On the **Julia/NUTS** path, though, `logcdf` is not
autodifferentiable for the discrete families, so a censored `poisson` or `negbin`
cell fails there. You have three options: run that fit on the **Stan** backend,
which uses analytic discrete CDFs and handles censored counts directly (this is
the one case where Stan is the strongest backend); put the cell on a continuous
(`gaussian` or `lognormal`) stream, as in the shipped `SIR` and `SEIR` examples;
or use the soft interval `[A,B]~s`, a plain penalty on the model mean
(`log g(mu)`, no CDF) that works for every family under NUTS. Discrete streams
can always carry plain observed and missing (`x`) cells on any backend.

The soft interval `[A,B]~s` says "most plausibly in `[A,B]`, tapering off outside
with scale `s`". At `A - s` (or `B + s`) the penalty is one log-unit. It is the
tool for "above 100, but most of the mass in 110–130": write `[110,130]~10`. A
hard `[110,130]` would instead forbid everything outside the band.

> **How censored and interval cells enter each fit.** In the MLE (least-squares)
> loss they become one-sided or two-sided hinge penalties: exactly zero while the
> model prediction satisfies the bound or sits inside `[A,B]`, then growing
> quadratically once it strays outside. In the Bayesian fit they are proper
> censored likelihood terms (`P(Y >= L)`, `P(A <= Y <= B)` via the CDF), which
> keep some curvature even inside the region. Note that `>L` versus `>=L` only
> matters for the discrete Bayesian CDF, not for the MLE hinge.

The `Likelihood` column declares each stream's family (Bayesian fits only; the
MLE loss is least-squares and ignores it). The families:

| Family | `Likelihood` value | Data | Notes |
|----|----|----|----|
| Gaussian | `gaussian` | continuous | proportional-error Normal; the default when the column is absent |
| Log-normal | `lognormal` | positive | |
| Poisson | `poisson` | counts | |
| Negative binomial | `negbin` | over-dispersed counts | mean--dispersion form, with a fitted dispersion per stream |

All four families work on all three backends. Each also has a couple of aliases
(`normal`, `lnorm`, `pois`, `nbinom`, `nb`), and non-integer data on a count
family is rounded, with a warning.

The column can also carry a global deviation for `A+`/`A-` cells:

```
gaussian                  # family only
gaussian; asym=4          # family plus a deviation of 4 (absolute units) for A+/A- cells
```

Using `A+`/`A-` in a stream whose `Likelihood` has no `asym=` is an error; there
is no silent default. Either add `; asym=<number>`, or use the explicit `A->B`
form, which carries its own deviation.

#### Under-reporting and coverage models

When "observed < true" comes from a mechanism, say you see only a fraction of
cases (a reporting rate, a cohort coverage), model it directly rather than as a
bound. The `Formula` column already refers to parameters by name, so give the
stream a coverage factor:

```
Formula       Parameters
rho*X2        rho=Beta(2,2)     # reported = rho * true infected; rho fitted
```

`rho` can be fixed (`*rho=0.7`) or fitted with any prior. This gets you more than
a bound does, but `rho` trades off with the epidemic scale, so it needs an anchor:
a second, fully observed stream, a fixed population `N0`, or an informative prior
on `rho`. This is a modelling pattern, not a cell type.

### dataDummy: derived series

These are extra series to evaluate on the trajectory and overlay on plots, handy
for looking at combinations of compartments (prevalence `X2/(X1+X2+X3)`, say, or a
subpopulation sum). Each row is just a `Label` and a `Formula` (the `Label` again
defaults to the `Formula`). Dummy series carry no likelihood and are never fitted.

## Simulate without fitting

If every state and parameter in the sheet is fixed (`*name=val`), there is nothing
to estimate. You can run the model forward with no `dataCombined` at all and
evaluate the dummy formulas on the trajectory:

```{r}
sim <- simulate_model(sc$modelParams, data_dummy = sc$dataDummy,
                      solver = solver_control(backend = "r"))
sim                          # compartments, time span, fixed parameter values
head(sim$sir_out)            # the solved trajectory
plot_simulation(sim)$grid    # one panel per compartment / evaluated series
```

If any quantity is still fittable, `simulate_model()` says so and points you at
`fitCompartmentalModel()`. Fixed-value `*_sim` example scenarios (`SIR_sim`,
`SEIR_sim`, and so on) ship with the package.

## Fit by maximum likelihood

```{r}
fit <- fitCompartmentalModel(
  modelParams  = sc$modelParams,
  dataCombined = sc$dataCombined,
  method       = "lbfgsb",                      # or "deoptim" (global), "hypercube"
  control      = optim_control(maxit = 1000),   # progress = FALSE for a silent fit
  solver       = solver_control()               # backend = "r" for a Julia-free fit
)

summary(fit)
get_point(fit)        # named point estimate (initial states plus parameters)
```

`method = "lbfgsb"` is a fast deterministic local search. `"deoptim"` is a
stochastic global search (differential evolution), more robust on rugged
objectives but slower. `"hypercube"` is a deterministic Sobol pre-scan. The
`*_control()` helpers bundle the settings for each stage; see `?optim_control`,
`?hypercube_control`, and `?solver_control`.

## Fit by Bayesian sampling

Pass `method = "bayes"` and a `bayes_control()` spec, and choose the sampler with
the `solver` backend (`"julia"`, `"stan"`, or `"r"`).

```{r}
fit_b <- fitCompartmentalModel(
  modelParams  = sc$modelParams,
  dataCombined = sc$dataCombined,
  method       = "bayes",
  solver       = solver_control(),                        # or backend = "stan" / "r"
  bayes        = bayes_control(chains = 4, iter = 2000, warmup = 1000)
)
```

`bayes_control(progress = FALSE)` silences the sampler's progress output on any
backend. Inspect the posterior and its convergence:

```{r}
posterior_report(fit_b)        # convergence verdict plus means and intervals
posterior_means(fit_b)
posterior_intervals(fit_b, p = 0.95)
plot_prior_posterior(fit_b)    # prior vs posterior, per parameter
identifiability_report(fit_b)  # narrowing, coverage, correlation diagnostics
```

A useful check is to compare the posterior against an MLE fit of the same model:

```{r}
fit_mle <- fitCompartmentalModel(sc$modelParams, sc$dataCombined, "lbfgsb")
compare_to_mle(fit_b, fit_mle) # posterior mean and median vs MLE
```

## Plot

`plot_fit()` returns per-stream `plots`, an arranged `grid` (if patchwork is
installed), and the editable plotting `code`. Overlay dummy series with
`data_dummy`.

```{r}
res <- plot_fit(fit_b, data_dummy = sc$dataDummy, band_type = "both")
save_grid(res, sc$plot_path)   # writes a new numbered file; never overwrites
```

`band_type` sets the uncertainty display for Bayesian fits: `"predictive"`,
`"mean"`, `"both"`, or `"spaghetti"`.

**Colour scheme.** Plots use the colour-blind-safe Okabe-Ito palette by default.
For print or monochrome output pass `palette = "grey"` (or `"grayscale"`), or a
custom palette list (see `?cfit_palette`):

```{r}
res_bw <- plot_fit(fit_b, palette = "grey")
```

To switch every plot at once, set the global option; a per-call `palette` still
overrides it:

```{r}
options(compfit.palette = "grey")   # back to colour: options(compfit.palette = "okabe")
```

## Save and reload

```{r}
save_fit(fit_b, "Saved fits/fit.rds")
fit_b <- reload_fit("Saved fits/fit.rds")   # re-registers the backend and rebuilds the loss
```

Use `reload_fit(path, register = FALSE)` for a results-only read (posterior draws,
no backend session).

## Sensitivity

Global (variance-based, Sobol) sensitivity of the model outputs, summarised:

```{r}
sob <- sobol_streams(fit_b, n = 500)
sobol_report(sob)              # validity, per-stream ranking, interactions, heatmap
```

Local (derivative) sensitivity at the fitted point:

```{r}
ls <- local_sensitivity(fit_b, type = "relative")
plot_local_sensitivity(ls, kind = "tornado")
```

## Reproducible export

`extract_code()` writes self-contained R source that reproduces the model, loss,
and/or fit from a fit object, handy for sharing or archiving an analysis:

```{r}
extract_code(fit_b, file = "reproduce.R", what = "all")
```

## Where next

- *"Fitting an SIR model with compfit"* is the narrated, run-live walkthrough.
- `?fitCompartmentalModel` covers the full fitting interface and its `method`s.
- `?solver_control`, `?optim_control`, `?bayes_control` document every option.
- `?simulate_model`, `?plot_fit`, `?plot_prior_posterior` cover simulating and
  plotting.
- `?sobol_report`, `?local_sensitivity`, `?identifiability_report` cover analysis.
