---
title: "Fitting an SIR model with compfit"
author: "Anna Bot"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,     # the R Bayes backend prints a loud caveat; surface it in prose instead
  warning  = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  dev = "svglite",      # vector SVG via svglite (no cairo/X11 needed); crisp on every device when hosted
  cache = TRUE          # heavy chunks (the fits) are cached; delete sir-tutorial_cache/ to force a rerun
)
library(compfit)
set.seed(1)
options(rmarkdown.html_vignette.check_title = FALSE)  # this is a pkgdown article, not a shipped vignette
```

## Introduction

Compartmental models are a fantastic tool to better understand how
epidemics or within-host dynamics of a virus unfold. Either you test
some hypothesis of how the dynamics change if you wiggle a parameter or
some initial conditions, or you want to figure out some unknown
parameters by fitting the model to actual data. If you have less than
ten compartments, there's usually no hassle in editing some details here
and there, be that in the model set-up or in the loss function used for
calibration. But, in fields like HIV research, where you need
stratifications according to clinical and behavioural markers, it
becomes cumbersome quite fast to go into your many lines of R code to
figure out where you need to change what. Not to mention when you want
to adapt a parameter to be time-varying as piecewise constant, or
linear, or any function, really.

This was the starting point for designing the package `compfit`. Instead
of maintaining (and struggling through) many lines of code and multiple
versions of the same model, it reduces the set-up to three csv or Excel
files, which encode the model structure, the parameters and states, as
well as any data used by the model. This way, one can cleanly iterate on
model versions and try out adaptations in a clean fashion.

### Who this package is for

This package was designed for a spectrum of users, with the two extremes
being 

- a user who knows basic R and has a compartmental model they want
to run or calibrate without the overhead of the entire machinery going
into it, and 
- an experienced user who wants the flexibility and ease
of the package but who also wants to fall back on the actual code if
need be. 

If there is any functionality you're missing, get in touch!

## The example: Susceptible-Infected-Recovered (SIR)

The core functionalities are exemplified using one of the most known
compartmental models, the SIR model. It is used to model the dynamics
and development of an epidemic, be that influenza, COVID-19, measles,
and many more. The population is split into the three subpopulations of
susceptible, infected and recovered individuals. A susceptible
indiviudal becomes infected when interacting with an infected
individual, and an infected individual is recovered after a certain
amount of time; these are the flows between the subpopulations. The
subpopulations are visually shown as boxes, or *compartments*, and the
flows between them are indicated by arrows, compare with the depiction
below.

```{r sir-diagram, echo = FALSE, out.width = "40%"}
# SIR diagram: a pre-rendered static SVG. Calling DiagrammeR::grViz() at render
# time (even piped to export_svg) registers an htmlwidget and breaks pkgdown's
# article build, so the SVG is committed instead. To change the graph, edit the
# DOT and regenerate with DiagrammeRsvg::export_svg(DiagrammeR::grViz(...)).
knitr::include_graphics("sir-diagram.svg")
```

To describe how the subpopulations change over time in size, we rewrite
this into ordinary differential equations (ODEs):

$$\begin{align}
\tfrac{dS}{dt} &= -\beta SI \\
\tfrac{dI}{dt} &= \beta SI - \gamma I \\
\tfrac{dR}{dt} &= \gamma I
\end{align}$$ The corresponding figure we designed before can capture
this information to some extent by adding the parameters governing the
flows above the arrow.

```{r sir-diagram-params, echo = FALSE, out.width = "40%"}
# SIR diagram with the flow parameters on the arrows: pre-rendered static SVG
# (see the note on the sir-diagram chunk above).
knitr::include_graphics("sir-diagram-params.svg")
```

Note that some of the parameters can intuitively be understood as
"moving an individual from one compartment to the next by this rate"
(i.e. the $\gamma$ term), but someone from the $S$ compartment is moved
to the $I$ compartment by way of a $\beta I$ term, involving the size of
the infected compartment. This makes complete epidemiological sense; but
mathematically it complicates solving this system of equations, and most
of the time we thus need numerical approaches, especially when the
compartmental model consists of many more compartments with multiple
infection terms.

You can imagine that this model cannot capture the dynamics of every
infectious disease. For HIV, for example, it makes sense to split it up
into susceptibles ($S$), acute infection ($A$), chronic infection ($C$),
diagnosed and under treatment ($T$). In that case, you also have many
more transmission terms between the susceptibles and all people living
with HIV; important to note is that after transmission, no matter which
compartment it was due, they always move to the compartment of acute
infections. The equations then may look like

$$\begin{align}
\tfrac{dS}{dt} &= -\beta_A SA-\beta_C SC - \beta_T ST \\
\tfrac{dA}{dt} &= \beta_A SA+\beta_C SC + \beta_T ST -\gamma_A A \\
\tfrac{dC}{dt} &= \gamma_A A -\gamma_C C \\
\tfrac{dT}{dt} &= \gamma_C C
\end{align}$$

You can extend this model in arbitrary ways, to fit the infectious
disease that you are interested in. To actually *solve* it
mathematically in R or Python, you need the values of the parameters
involved as well as the sizes of the compartments at time $0$. Usually,
one sets up a pipeline to either 

1. see how the model develops for
different values (meaning one knows all the parameters and initial
values), or 
2. figure out the parameters and initial values needed to
explain epidemiological data, which we call **model fitting** from now
on. 

The second objective is very common when you know part of the
parameters and initial values, and want the model to best describe the
data you have. For the HIV example, one can assume a certain value for
the infection terms; indeed, under successful treatment, $\beta_T = 0$.
We can also say that the acute infection is over after a fairly fixed
amount of time, giving $\gamma_A$. But it is difficult to know the
diagnosis and treatment rate $\gamma_C$.

For both objectives, you'd set up the equations in a model function, and
then solve it using `ode()` from the package `deSolve`, or by a similar
method. For model fitting, you'd then set up a loss function based on
this compartmental model function and run `optim()` over it for a
maximum likelihood estimation (MLE). If instead one leans Bayesian, then
one sets priors on the unknown parameters and initial states, as well as
the likelihood families for the data one has, and then builds a Stan
model or similar.

This is heavy machinery. And heavy machinery means multiple lines of
code to maintain, which is, as said before, very cumbersome. So instead,
below, based on the SIR model, I will outline the functionalities of
`compfit` that bundle up what one wants from simulation, or MLE/Bayes
model fitting.

## Encoding the set-up

First, install and load `compfit`.

```{r library, eval = FALSE}
# install.packages("remotes")
remotes::install_github("OWNER/compfit")
library(compfit)
```

The input for the functions of `compfit` consists of three csv files (or
Excel): 1. The model structure, parameters and states encoded in
**modelParams.csv**. 2. The data points used for model fitting and
plotting in **dataCombined.csv** (which is optional if one only wants to
simulate!). 3. The dummy compartments used for inspecting the output of
either a simulation or a model fit, in **dataDummy.csv**. The three
files should be stored in the same folder, which we call "Scenario". We
read it in using the custom `load_scenario()` function.

```{r load-scenario}
sir_dir <- system.file("extdata", "SIR", package = "compfit")

sc <- load_scenario(
  sir_dir,
  combined_file = "dataCombined.csv",
  dummy_file    = "dataDummy.csv",
  params_file   = "modelParams.csv"
)
```

Using the read-in files, we discuss what needs (and what can) go into
the model files

### Model structure

The file consists of required columns, as well as optional columns.

#### Required

- States
- Parameters
- Constant, Linear\<i\>, Quadratic\<i\>
- Others

```{r show-modelParams-required, results='asis', echo = FALSE}
options(knitr.kable.NA = "")   # show blank cells as blank rather than "NA"
req_cols <- c("States", "Parameters",
              "Linear1", "Quadratic1", "Linear2", "Quadratic2",
              "Linear3", "Quadratic3", "Others")
tbl <- knitr::kable(sc$modelParams[, req_cols], format = "html")  # escapes <, >, &
tbl <- gsub("*", "&ast;", tbl, fixed = TRUE)   # literal * -> entity, so Markdown doesn't read it as italics
# nowrap keeps entries like -(gamma_t) on one line; the scroll box handles the width
cat('<style>.cf-modeltable td, .cf-modeltable th { white-space: nowrap; }</style>')
cat('<div class="cf-modeltable" style="overflow-x:auto">', tbl, '</div>', sep = "")
```

The **States** column encodes the compartments of the model. In this
example we opted for calling them X1, X2, X3 corresponding to S, I, R,
but you could equally call them S, I, R directly. To keep it well
organised, we rewrite the SIR model equations using the new names for
the states:

$$\begin{align}
\tfrac{dX_1}{dt} &= -\beta X_1X_2 \\
\tfrac{dX_2}{dt} &= \beta X_1X_2 - \gamma X_2 \\
\tfrac{dX_3}{dt} &= \gamma X_2
\end{align}$$

The **Parameters** column encodes the parameters used in the model.
Every state and parameter need a value; for the state, this corresponds
to an initial value at time $0$. If one writes "X3=0", then this means
that it's a fixed value and thus not used for model fitting. The
asterisk is mandatory for fixed values (but I'm open for changes here).
In this case, this means that at the beginning of an epidemic, we have
no recovered individuals.

Writing "beta=[0,3]" encodes the fact that we do not know $\beta$'s true
value but we believe it to lie somewhere in the interval from 0 to 3 (if
it's a Bayesian framework, then this represents the uniform prior on
that interval). The starting point for that parameter will be the midpoint of the interval; if one prefers to start from somewhere else, the syntax is "beta=[0,3]|2". The set-up is flexible enough to accommodate for us not
knowing how many individuals were infected at the start, thus the
infected compartment having the shape "X2=init_inf_0\*N0_0", and the
corresponding shape for the susceptible compartment.

One can also choose different priors on the parameters and initial states. Here are the options with their syntax:

| Prior | Sheet syntax | Support |
|----|----|----|
| Uniform | `[lo,hi]` | `[lo,hi]` |
| Normal | `Normal(mu,sd)` | real line |
| Log-normal | `LogNormal(mu,sd)` | positive |
| Beta | `Beta(a,b)` | `(0,1)` |
| Gamma | `Gamma(a,b)` | positive |
| Student-t | `StudentT(nu,mu,sigma)` | real line (heavy tails; `nu` = df) |

Some parameters can appear in the form of functions; indeed, $\gamma_t$
is an extension of the $\gamma$ term we have in the model. It is entered
as "gamma_t\<-gamma\*(1+ramp\*time)" under **Functions**, which we
discuss below. The gamma in there is estimated, and thus must appear
under **Parameters**. The new model looks like:

$$\begin{align}
\tfrac{dX_1}{dt} &= -\beta X_1X_2 \\
\tfrac{dX_2}{dt} &= \beta X_1X_2 - \gamma_t X_2 \\
\tfrac{dX_3}{dt} &= \gamma_t X_2
\end{align}$$ where $\gamma_t$ is a function in time.

Since the SIR model has three compartments, we have columns **Linear1,
Quadratic1, Linear2, Quadratic2, Linear3, Quadratic3**. Any pair
Linear\<i\>, Quadratic\<i\> encodes the equation describing that
compartment's inflow and outflow. For $X_1$, which is the compartment S,
we only have the outflow $-\beta SI = -\beta X_1 X_2$, which is a
quadratic term. Thus, the Linear1 column can be left empty. In the
Quadratic1 column, we go to the second row, since the $3(i-1) + j$-th
row in corresponds to the parameter in front of the term $X_iX_j$. So the entry "-beta" encodes the fact that $-\beta X_1 X_2$ is the quadratic term in $\tfrac{dX_1}{dt}$. The number between asterisks before "-beta" tells us to where this flows to, which for this model may not look very relevant, but as discussed before, it comes up in an HIV-type compartmental model. To see how
Linear\<i\> works, look at Linear3. In the second row, we have the entry
"(gamma_t)", corresponding to the linear term $\gamma_t X_2$ in the model
equations.

*Note*: one can only fill in the non-zero terms and leave the rest
empty, or fill it with 0's for book-keeping. In fact, one can even omit an entire column, which then translates to no terms of that type added to the equations; in our example, one could entirely omit Quadratic2 and Quadratic3. However, no error or
warning is shown if one writes the terms in the wrong line or erroneously omits a column. 

Not appearing in the model is the **Constant** column, where the first row adds any constant term to $\tfrac{dX_1}{dt}$. Examples for such terms would be birth or migration terms.

Lastly, **Others** is a required column. One has to provide: 

- *startpoint*: corresponding to time 0; it has to be an integer, usually a
year. *Note the convention chosen here that startpoint - 1 is the time
for the initial values for the states!* In this file, we start in 2010,
and the initial states are from 2009. 
- *endpoint*: until when one has
data or wants to end the simulation, strictly larger than startpoint and
the same unit. In the provided file that's 2018. 
- *partition*: the number
into which we divide one integer; partition = 365 means we divide a year
into days, partition = 12 means the division is by months.

#### Optional

- \_Level\<i\>
- Functions
- Conditions

```{r show-modelParams-optional, results='asis', echo = FALSE}
opt_cols <- c("_Level1", "Functions", "Conditions")
opt <- sc$modelParams[, opt_cols]
# keep only rows that actually hold something (drops the empty trailing rows)
opt <- opt[apply(opt, 1, function(r) any(!is.na(r) & trimws(as.character(r)) != "")), ]
tbl <- knitr::kable(opt, format = "html")
tbl <- gsub("*", "&ast;", tbl, fixed = TRUE)   # literal * -> entity (Functions cell has gamma*(1+ramp*time))
cat(tbl, sep = "")
```

**\_Level1** in our example is optional (but the underscore before the name isn't!), and it encodes which compartment belongs to which sublevel. A typical sublevel divison would be by age, or by biological sex, if there is no uniform mixing. So if we had 2 levels, one would need to specify to which \_Level1 or \_Level2 a
compartment belongs, either by its corresponding number or by its name.

We already implicated how **Functions** is used: and indeed, the gamma_t
is used in the model structure equations. The way gamma_t is entered is
using the R language "\<-". Lastly, one may want to enforce conditions
on some of the parameters and initial states being fixed, which can be
done in **Conditions**.

### Data points

The data streams are optional if one only wants to simulate and not
model fit. The file is organised by **Label**, which is what is
displayed later in the plots for easy referencing, **Formula** which
encodes what the label describes, **Likelihood** describes the
likelihood family we ascribe to the data stream, **Weight** represents
the trust we have in the data source (1 is normal, 2-3 is strong, 4-7 is
very strong), **Average** is used when doing MLE to bring all data
streams to a common magnitude so the solver does not prioritise the
streams which have the relatively largest numbers. **Average** is optional, and in its absence the data stream is automatically weighted by the inverse of the data stream mean. **Weight** is also optional and defaults to 1 on all nonspecified data streams.

Note that **Weight** and **Average** are only used in the MLE fit, whereas **Likelihood** is only used in a Bayesian fit.

```{r show-data_dataCombined, results='asis', echo = FALSE}
tbl <- knitr::kable(sc$dataCombined, format = "html")
tbl <- gsub("*", "&ast;", tbl, fixed = TRUE)   # literal * -> entity, so Markdown doesn't read it as italics
cat('<style>.cf-modeltable td, .cf-modeltable th { white-space: nowrap; }</style>')
cat('<div class="cf-modeltable" style="overflow-x:auto">', tbl, '</div>', sep = "")
```

Then, one needs to specify the data for each year. Entering a number means
the model fits to that precise number. If one sets [15,95] as for
"Recovered (X3)" in 2012, it means that the true value is somewhere
between 15 and 95. And if one writes -- as was done in 2018 -- that
\>=2147, then this encodes the knowledge that the number is at least
2147. There are more options:

| Cell | Meaning |
|----|----|
| a number | observed value (`0` is a genuine zero) |
| `x` or blank | missing / unobserved |
| `<L` / `<=L` | left-censored: true value strictly / at-or-below `L` |
| `>L` / `>=L` | right-censored: true value strictly / at-or-above `L` |
| `[A,B]` | interval-censored: true value in `[A,B]`, flat within (`A < B`) |
| `[A,B]~s` | **soft** interval: plateau `[A,B]`, exponential shoulders of scale `s` (`~sl,su` for asymmetric shoulders) |
| `A->B` | asymmetric: recorded `A`, soft toward `B` (deviation `\|B-A\|`) |
| `A+` / `A-` | asymmetric: recorded `A`, soft up/down, deviation from the stream's `Likelihood` column |

### Dummy compartments

Lastly, we have the dummy compartments, which are useful both for
diagnostics when model fitting, but equally useful when examining
certain combinations of subpopulations -- this is for "opening the black
box". The structure is a simpler version of the data streams encoding,
needing just **Label** and **Formula**.

```{r show-data_dataDummy, echo = FALSE}
knitr::kable(sc$dataDummy)
```


## Fit by maximum likelihood

In a maximum likelihood estimation (MLE, done via least squares), we let a solver
deterministically move inside the parameter space to continually improve the parameter choice. That is why we set method = "lbfgsb" (L-BFGS stands
for Limited memory Broyden-Fletcher-Goldfarb-Shanno), which is a quasi-Newton solver; a different option is "deoptim", where the solver is started multiple times from different
starting points, and should in principle avoid local optima and find a
global one. Usually, "lbfgsb" works fine. Of course, the time it takes
to model fit scales with the number of compartments, and the number of
parameters and states being model fit.

```{r fit-mle}
fit_mle <- fitCompartmentalModel(
    sc$modelParams, sc$dataCombined,
    method = "lbfgsb",
    solver = solver_control(backend = "r"),
  control = optim_control(progress = FALSE)
  )

get_point(fit_mle)
```

The `get_point(fit_mle)` prints the estimations. Using `plot_fit()`, we can look at how well the solver fit to the data, and inspect the dummy variables too.

```{r plot-mle}
plot_fit(fit_mle, data_dummy = sc$dataDummy)$grid
```

We see that with the fitted parameters, the predicted curves fit nicely through the data points. In principle, one could replace all the parameters like "beta=[0,3]" with "beta=[0,3]|random", which draws uniformly from the interval to start the estimation. Then, one can launch it multiple times, getting a proxy for a confidence interval. The more natural approach, though, is a Bayesian set-up.

## Fit by Bayesian sampling

If one prefers a Bayesian set-up (and is willing to wait a lot longer than for an MLE fit), `compfit` provides an `R`, `Julia` or a `Stan` backend to fit the model instead. This is controlled by the solver, with options: `backend="r"`, `backend="julia"` or `backend="stan"`. The `R` backend uses the gradient-free sampler `BayesianTools`, whereas the Julia path needs a bridge to Julia and uses a No U-Turn sampler (NUTS), which has to first be initiated. Stan also uses a NUTS. In fact, the `R` backend is only sensible for really small models but it is slow and mixes poorly; we illustrate this below and compare it to Julia (and then Stan), so one should understand the output from the `R` backend as a cautionary tale. 

After setting up the files and reading them in via `load_scenario`, we can call `fitCompartmentalModel` by providing the model structure, the data points as well as some other controls. The main one is setting method to "bayes", and choosing the backend = "r". In bayes_control, choose the number of iterations and the number of chains. Whenever there is a progress option, it defaults to TRUE and you can see how the solver progresses.

```{r fit-bayes}
fit_bayes <- fitCompartmentalModel(
  sc$modelParams, sc$dataCombined,
  method  = "bayes",
  solver  = solver_control(backend = "r"),
  bayes   = bayes_control(iter = 4000, chains = 3, progress = FALSE)
) 
```

Once the fit is through, we can inspect how well it worked by looking at the output of `posterior_report()`:

``` {r posterior-report-r}
posterior_report(fit_bayes)
```

Well, clearly something is off if ESS is zero. We can look at how the posteriors of the fitted quantities look like using `plot_prior_posterior()`, and again we see it looks a bit off, though each posterior narrowed and moved away from the uninformative uniform prior. 

```{r plot-prior-posterior, fig.width = 6, fig.height = 3.5, out.width = "70%"}
plot_prior_posterior(fit_bayes)
```

At this point, it looks strange, but we persist still and compare the Bayes median and mean of the parameter posteriors to what we obtained from the MLE:

```{r compare}
compare_to_mle(fit_bayes, fit_mle) 
```

The MLE and the Bayes fit are not too far off from each other, which is encouraging. Since Bayes gives us directly credibility intervals, we also inspect that using `plot_fit()`, where the darker shade is coming from the parameter uncertainty, and the lighter one from the likelihood families chosen.

```{r plot-bayes}
plot_fit(fit_bayes, data_dummy = sc$dataDummy, band_type = "both")$grid
```

Another view of the same fit is a *spaghetti plot*, which draws one line per
posterior draw instead of a summarised band:

```{r plot-bayes-spaghetti}
plot_fit(fit_bayes, data_dummy = sc$dataDummy,
         band_type = "spaghetti", spaghetti_draws = 80)$grid
```

Now this is really weird, with this huge CrI and only very few curves from the spaghetti plot.

## Why the R-backend posterior looks degenerate

The issue is what we mentioned earlier: the R backend uses a gradient-free sampler and thus barely moves (in fact, it has a really hard time with the >=2147 and [15,95] information on the data streams, which leads it to reject a lot of the proposals and makes it virtually immobile). We can highlight this by inspecting how many different draws there actually are:

```{r unique-draws}
draws <- posterior_draws(fit_bayes)
nrow(draws)
nrow(unique(round(as.data.frame(draws), 6)))
```

Though there are 6006 draws, there are only 6 unique ones. Clearly, we need to move on to a better sampler.

## The same model with the Julia backend

Julia, which employs the No U-Turn sampler (NUTS) is one option to get around the
mixing problem we observed. Installing Julia on a build server is
heavy, so this article does **not** run it live: the fit is produced once and
saved, then loaded here for display. 

```{r julia-setup-reference, eval = FALSE}
# The number of threads have to be set before, and once they are set, 
# you cannot change it anymore in the session
Sys.setenv(JULIA_NUM_THREADS = 4)

library(compfit)
setup_julia()   
# starts Julia; the first call installs the Julia packages
                # (OrdinaryDiffEq, Turing, Distributions) and takes a few minutes
```

```{r julia-fit-reference, eval = FALSE}
# The runnable version of this lives in
# data-raw/make_julia_artifacts.R.
fit_julia <- fitCompartmentalModel(
  sc$modelParams, sc$dataCombined,
  method = "bayes",
  solver  = solver_control(backend = "julia"),
  bayes  = bayes_control(chains = 4, iter = 2000)
)
save_fit(fit_julia, "vignettes/articles/julia_fit.rds")
```

```{r julia-detect, include = FALSE}
# Resolve saved artifacts whether knitting from the article dir or the package root.
.find <- function(f) if (file.exists(f)) f else file.path("vignettes/articles", f)
julia_fit_path   <- .find("julia_fit.rds")
have_julia_fit   <- file.exists(julia_fit_path)
julia_bands_path <- .find("julia_bands.svg")       # predictive + mean CrI bands
have_julia_bands <- file.exists(julia_bands_path)
julia_spag_path  <- .find("julia_spaghetti.svg")   # per-draw trajectories
have_julia_spag  <- file.exists(julia_spag_path)
```

```{r julia-report, eval = have_julia_fit}
fit_julia <- reload_fit(julia_fit_path, register = FALSE)
posterior_report(fit_julia)
```

Now this looks much better, and the prior versus the posteriors of the parameter distributions also looks much more encouraging:

```{r julia-prior-post, eval = have_julia_fit, fig.width = 6, fig.height = 3.5, out.width = "70%"}
plot_prior_posterior(fit_julia)
```

```{r julia-bands-img, echo = FALSE, eval = have_julia_bands, out.width = "80%"}
# Pre-rendered on the Julia machine (plot_fit re-solves the ODE, so it needs a
# live Julia session and cannot be produced from a register = FALSE reload).
knitr::include_graphics(julia_bands_path)
```

The same can be said for the spaghetti plots, now with many more curves accepted.

```{r julia-spaghetti-img, echo = FALSE, eval = have_julia_spag, out.width = "80%"}
# Pre-rendered on the Julia machine (same reason as the band plot above).
knitr::include_graphics(julia_spag_path)
```

Julia is thus a much more robust option. For anyone who prefers Stan instead, we can see the same results as with Julia in the next section.

## The Stan backend

If installing Julia is a barrier, `compfit` can also sample with **Stan** via
`rstan` — the same gradient-based NUTS, no Julia required (it compiles a Stan
program on first use). Stan also mixes well, as one can see by again inspecting the different parts of the model, as done below.

```{r stan-reference}
fit_stan <- fitCompartmentalModel(
  sc$modelParams, sc$dataCombined,
  method = "bayes",
  solver = solver_control(backend = "stan"),
  bayes  = bayes_control(chains = 4, iter = 2000, progress = FALSE)
)
posterior_report(fit_stan)
```


```{r plot-bayes-stan}
plot_fit(fit_stan, data_dummy = sc$dataDummy, band_type = "both")$grid
```


```{r plot-bayes-spaghetti-stan}
plot_fit(fit_stan, data_dummy = sc$dataDummy,
         band_type = "spaghetti", spaghetti_draws = 80)$grid
```



## Coming soon: counterfactual scenarios and predictions

Once one has fitted a model and is happy with it, oftentimes it is of
scientific value to ask "what if ... ?" This can be encoded in
counterfactual scenarios and will be part of a future iteration of the
package. Similarly, a fitted model can serve to have some predictions
for the dynamics of the model as it goes beyond the last time point.
This is also in the works.

## Where next

- `vignette("compfit")` — the full workflow and the model-sheet grammar.
- `?fitCompartmentalModel`, `?plot_fit`, `?plot_prior_posterior` — the
  reference pages.
